---
title: "Food content check"
output: html_notebook
---

We start by reading the extracted food contents.

```{r}
library(data.table)

contents <- fread("zcat < ~/projects/2024_medi_paper/db/data/dbs/food_contents.csv.gz")
head(contents)
```

We will now check a few main properties.

## Energy content

The food contents contain measured energy content extracted from the FOODB which may be several measurements per food
items and component. We also include energy content calculated by the Adwater method for each food and preparation type.
Let's extract those now.

```{r}
en <- contents[grepl("Energy", name)]
wide <- dcast(
  en, 
  preparation_type + wikipedia_id + food_id ~ name, 
  value.var="standard_content", 
  fun.aggregate=mean
)
wide
```

No we will compare those:

```{r}
library(ggplot2)
theme_minimal() |> theme_set()
wide[preparation_type %in% c("", "no"), preparation_type := "unknown"]
wide[, cor.test(Energy, `Energy (calculated)`)]

ggplot(wide) +
  aes(x=Energy, y=`Energy (calculated)`, color=tolower(preparation_type)) +
  geom_abline(slope=1, linewidth=0.8) +
  geom_point()
```

## Cholesterol adjustments

It is likely that for a subset of measurements cholesterol abundance was incorrectly transformed from ug to mg. This can be observed when looking at the
histogram of the original values:

```{r}
ggplot(contents[name == "Cholesterol"]) +
  aes(x=orig_content) +
  geom_histogram(bins=50) +
  scale_x_log10()
```

Given that even very high cholesterol foods (eggs, bacon etc) never exceed a few 100mg/100g the second peak is likely affected by the conversion
error. Those values were converted to the correct mg amounts:

```{r}
ggplot(contents[name == "Cholesterol"]) +
  aes(x=standard_content) +
  geom_histogram(bins=50) +
  scale_x_log10()
```

We can also check for some common food items and their cholesterol content:

### Chicken (around [64-111 mg/100g](https://www.medicalnewstoday.com/articles/cholesterol-in-chicken#by-part-of-the-chicken))

```{r}
contents[name == "Cholesterol" & wikipedia_id == "Chicken", summary(standard_content)]
```

### Beef (around [78-389 mg/100g](https://www.ucsfhealth.org/education/cholesterol-content-of-foods))

```{r}
contents[name == "Cholesterol" & wikipedia_id == "Cattle", summary(standard_content)]
```

### Cod (around [43 mg/100g](https://fdc.nal.usda.gov/fdc-app.html#/food-details/171955/nutrients))

```{r}
contents[name == "Cholesterol" & wikipedia_id == "Pacific_cod", summary(standard_content)]
```
