# Food content validaton PATH1

```{r}
library(data.table)
library(ggplot2)
theme_minimal() |> theme_set()
```

We start by reading in the general metadata to get our sequencing to individual mapping.

```{r}
meta <- fread("data/metadata.csv")
meta[, "sample_id" := tstrsplit(MetagenomicsFileR1, "_R")[[1]]]
meta[, "Participant ID" := sprintf("PA%03d", SubjectID)]
meta
```

We will also extract the FFQs.

```{r}
vars <- names(meta)
ffq_vars <- vars[grepl("D_FFQ", vars)]
idvars <- c(vars[!vars %in% ffq_vars], "D_FFQ_AV_GRAMWT_G_USDA")
ffq <- melt(meta, id.vars=idvars, value.name="ffq_amount", variable.name="name")[ffq_amount > 0]
ffq[, ffq_amount := as.numeric(ffq_amount) / as.numeric(D_FFQ_AV_GRAMWT_G_USDA)]
head(ffq)
```

And we read the poop collection dates and merge.

```{r}
dates <- fread("data/collection_dates.csv")
dates[, "time" := as.POSIXct(paste(Date, Time), format="%m/%d/%Y %I:%M %p")]
dates[, "SubjectID" := tstrsplit(ID, "-")[[2]] |> as.numeric()]
dates[, "Period" := tstrsplit(ID, "-")[[3]] |> as.numeric()]
merged <- dates[, .(SubjectID, Period, time)][meta, on=c("SubjectID", "Period")]
merged
```

Now we read the 24h intake data.

```{r}
scales <- c(
  kcal = 100,
  IU = 100,
  g = 100 * 1000,
  mg = 100,
  mcg = 100 / 1000
)

intake <- fread("data/P4W1204.txt")
idvars <- c(names(intake)[1:19])
intake <- melt(intake, id.vars=idvars, variable.name = "measure", value.name = "amount")[amount > 1e-6]
intake[, amount := as.numeric(amount)]
intake <- intake[merged[, .(sample_id, `Participant ID`, Period, time, Treatment, BaselineEnd)], on="Participant ID", nomatch=0]
intake[, amount := amount / (`Total Grams`)]
intake[, original_unit := stringr::str_match(measure, "\\((\\w+)\\)$")[,2]]
intake[, amount := amount * scales[original_unit]]
intake[, "entry" := as.POSIXct(`Date of Intake`, format="%m/%d/%Y")]
intake[, "time_diff" := time - entry]
fwrite(intake[!is.na(original_unit)], "data/intake_processed.csv")
intake
```

Now we read the MEDI abundances for food content.

```{r}
medi <- fread("data/food_content.csv")
medi[name == "Cholesterol", abundance := abundance]
medi
```

Let's compare the macronutrients. For this we will map the names for the major nutrients.

```{r}
recall_map <- c(
  Energy = "Energy (kcal)",
  Proteins = "Total Protein (g)",
  `Fat` = "Total Fat (g)",
  Carbohydrate = "Total Carbohydrate (g)",
  `Fiber (dietary)` = "Total Dietary Fiber (g)",
  Cholesterol = "Cholesterol (mg)",
  Cobalamin = "Vitamin B-12 (cobalamin) (mcg)",
  `Potassium` = "Potassium (mg)"
)

ffq_scales <- c(
  `Energy (kcal)` = 100,              # kcal per 100g
  `Total Protein (g)` = 100 * 1000,  # mg per 100g
  `Total Fat (g)` = 100 * 1000,
  `Total Carbohydrate (g)` = 100 * 1000,
  `Total Dietary Fiber (g)` = 100 * 1000,
  `Cholesterol (mg)` = 100,
  `Vitamin B-12 (cobalamin) (mcg)` = 100 / 1000,
  `Potassium (mg)` = 100
)

ffq_map <- c(
  D_FFQ_AV_ENERGY_KCAL_USDA = "Energy (kcal)",
  D_FFQ_AV_TOTAL_PROTEIN_G_NDSR = "Total Protein (g)",
  D_FFQ_AV_TOTAL_FAT_G_USDA = "Total Fat (g)",
  D_FFQ_AV_CARBOHYDRATE_G_USDA = "Total Carbohydrate (g)",
  D_FFQ_AV_TOTAL_DIETARY_FIBER_G_NDSR = "Total Dietary Fiber (g)",
  D_FFQ_AV_CHOLESTEROL_MG_USDA = "Cholesterol (mg)",
  D_FFQ_AV_Supp_Vitamin_B12 = "Vitamin B-12 (cobalamin) (mcg)",
  D_FFQ_AV_POTASSIUM_MG_USDA = "Potassium (mg)"
)

medi[, "measure" := recall_map[name]]
ffq[, "measure" := ffq_map[name]]
comparison <- intake[medi, on=c("sample_id", "measure"), nomatch=0]
ffq_comparison <- ffq[!is.na(measure)][medi, on=c("sample_id", "measure"), nomatch=0, allow.cartesian=T]
comparison <- comparison[BaselineEnd == "End"]

#comparison[, amount := amount * scales[measure]]
ffq_comparison[, ffq_amount := ffq_amount * ffq_scales[measure]]
```

```{r}
ggplot(comparison[BaselineEnd == "End"]) +
  aes(x=time_diff) +
  geom_histogram(bins=20, fill="white", color="black")
```

```{r}
means <- comparison[time_diff>48, .(
  amount=mean(amount), 
  sd=sd(amount),
  se=sd(amount)/sqrt(.N),
  time=time[1], 
  n=uniqueN(entry), 
  subject=`Participant ID`[1],
  diet=Treatment[1]
  ), by=c("measure", "sample_id", "abundance")]

tests <- comparison[time_diff>48, cor.test(amount, abundance, method="pearson")[c("p.value", "estimate")], by="measure"]
tests
```

```{r, fig.width=10, fig.height=4}
ggplot(means) +
  aes(x=amount, y=abundance) +
  geom_pointrange(aes(xmin=amount-se, xmax=amount+se)) +
  geom_point() +
  stat_smooth(method="lm") +
  facet_wrap(~ measure, scales="free", nrow=2) + 
  geom_text(data=tests, aes(x=Inf, y=Inf, label=sprintf("r=%.2g, p=%.2g", estimate, p.value)), size=3.5, vjust=1, hjust=1) +
  labs(x="consumed [mg/100g]", y="detected in stool [mg/100g]") + scale_y_continuous(expand=c(0.25, 0))
ggsave("nutrients_48h.pdf", width=10, height=4)
```



FFQ comparison

```{r, fig.width=6, fig.height=4}
ggplot(ffq_comparison) +
  aes(x=ffq_amount, y=abundance) +
  geom_point() +
  stat_smooth(method="lm") +
  facet_wrap(~ measure, scale="free")

ffq_comparison[, cor.test(ffq_amount, abundance, method="pearson")[c("p.value", "estimate")], by=c("measure")]
```
By offset groups:

```{r, fig.width=9, fig.height=2}
grouped <- comparison[BaselineEnd=="End"]
grouped[, "time_group" := Hmisc::cut2(as.numeric(time_diff), c(0, 48, 96))]
for (m in grouped[, unique(measure)]) {
  pl <- ggplot(grouped[measure == m]) +
    aes(x=amount, y=abundance) +
    geom_point() +
    stat_smooth(method="lm") +
    facet_wrap(~ time_group, scales="free", nrow=1) +
    ggtitle(m)
  show(pl)
}
```

```{r, fig.width=6, fig.height=3}
stats <- grouped[, .(r = cor(amount, abundance), n=.N), by=c("measure", "time_group")]
ggplot(stats) +
  aes(x=time_group, y=r, group=measure, color=measure) +
  geom_hline(yintercept=0, linetype="dashed") +
  geom_line() + geom_point() +
  labs(x="Δt stool after diet assessment [days]", y="Pearson r [measured vs. MEDI]", color="Nutrient")
ggsave("../figures/path_offset_stats.pdf", width=6, height=3)  
```

## Figure Data

We assemble all the tables first

```{r}
fig_data <- list(
  `Fig. 3B` = fread("../figure_data/mbd_nmds.csv"),
  `Fig. 3C` = fread("../figure_data/mbd_total_food.csv"),
  `Fig. 3D` = fread("../figure_data/path1_food_tests.csv"),
  `Fig. 3E` = means
)

si_figs <- list(
  `Fig. S2A` = fread("../figure_data/mbd_food_counts.csv"),
  `Fig. S2B` = stats
)
```

And we save them.

```{r}
openxlsx::write.xlsx(fig_data, "../figure_data/fig3.xlsx")
openxlsx::write.xlsx(si_figs, "../figure_data/figS2.xlsx")
```